Performance analysis of high-resolution \\ ice sheet simulations

### 2022-10-06 paper/perfmod.tex
ongoing edits

\newcommand{\nn}{\text{n}}
A diffusive description of the mass continuity equation is also valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ in the isothermal case, where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.}  While Equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.  Generically across stress balance choices, for grounded glaciers and ice sheets one observes that large values of $d$ indicate locations of unstable mode growth if explicit time-steps are chosen too large.

@@

Table \ref{tab:notation} lists the parameters used in our performance analysis.  The primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.  \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits, and big-O notation is only used in this limit.

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry-update (and velocity) solution requires $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit, coupled Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of map-plane domain & km \\
$m$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step in model years & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis.  Note $\alpha$, $\beta$, $\gamma$, and $m$ are pure numbers.}
\label{tab:notation}
\end{table*}

If the map-plane model domain is of width $L$ then these primary parameters are related by
\begin{equation}
\Delta x = O\left(\frac{L}{\sqrt{m}}\right) \quad \text{and} \quad m = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
\end{equation}
For a given domain width $L$ there are $O(\sqrt{m})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$.  However, our analysis addresses spatially-3D models with map-plane horizontal meshes.}

A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.  Storing these model state variables, plus the thermodynamical state, requires $O(m)$ memory if the mesh/grid has \emph{a priori} bounded resolution in the vertical direction.  The amount of fast memory needed by the simulation is also $O(m)$ if prior states are discarded or transferred to storage.  Such ice sheet models also have $O(m)$ velocity variables, but these are not state variables.  That is, a very-viscous stress balance computes velocity as a function of the true state variables.

Our assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,BrinkerhoffJohnson2015,Hoffmanetal2018,Lengetal2012,
Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.  Some solvers use 3D refinement \citep{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016}, with various distinctions between how horizontal and vertical meshing is handled.  This and many other details cannot be pursued here.

Ice sheet models resolve climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle, and longer scales.  Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.  Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.  Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ describes the frequency on which ice geometry is updated using ice velocity, i.e.~via solution of the mass continuity or surface kinematical equation.

Current-technology ice sheet models use explicit time-stepping which is only conditionally stable.  For the spatial resolutions used in present-day scientific applications, we observe that maintenance of explicit time-stepping stability requires steps substantially shorter than $1/q$ model years.

For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value, i.e.~of the $d$ in equation \eqref{eq:siamasscontinuity}.  For Stokes or other membrane-stress-resolving dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, represents the \emph{optimistic} paradigm.  The corresponding \emph{pessimistic} paradigm requires $\Delta t < O(D^{-1} \Delta x^2)$, using a representative diffusivity value $D$ computed in the SIA manner.

Explicit time-stepping with hybrid and higher-order schemes is somewhat better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.  Some hybrid schemes apply the pessimistic paradigm as an adaptive restriction \citep{Winkelmannetal2011}.  Other models apparently require the user to choose a fixed time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.  The optimistic paradigm is supported in theory for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).

Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$.  This restriction reflects the simulation purpose, not maintenance of stability.  For an implicit scheme the frequency of climate coupling determines the total simulation cost, according to the (large) solution cost, at each time step, of coupled mass and momentum equations.

For each explicit time step of a model using a membrane-stress-resolving balance the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.  For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 1$ if sparsity is exploited or $\alpha \approx 2$ if not \citep{Bueler2021}.

@@

By contrast, a multigrid method \citep{Trottenbergetal2001} can greatly reduce $\alpha$.  For example, Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[Table 8.1]{IsaacStadlerGhattas2015}.  The \cite{Lengetal2012} algebraic-multigrid Stokes solver may have similar scaling, but reported results do not constrain $\alpha$.  For a higher-order stress balance on the Greenland ice sheet, \cite{Tuminaroetal2016} report total algebraic-multigrid-preconditioned iterations suggesting $\alpha \approx 0.05$,\footnote{See Table 7.5.  Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.} and a geometric multigrid method by \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.

Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling or solver complexity \citep{Bueler2021}.  At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula; an SIA velocity solution therefore requires optimal $O(m)$ flops.

When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, strongly depending on solver design.  Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling.  Actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.  For example, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results, among others, all show good parallel scaling, something we do not address here.

Regardless of the stress balance, an explicit time-stepping scheme then applies the mass continuity equation to update the ice thickness using $O(m)$ work.  That is, once the velocity is computed for the previous time-step's geometry, an explicit scheme replaces old thickness values by new ones using a pointwise formula.  Any additional computation needed to remesh the updated geometry, a design-dependent cost, is omitted here.

@@

Now, how many flops are needed to simulate one model year?  Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ steps per model year.  We may write stability restrictions as required numbers of steps per model year.  That is, $\Delta t^{-1}$ is bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.  Recalling scaling \eqref{eq:paramrelation} of $\Delta x$ with $m$, in the explicit SIA and pessimistic-Stokes cases we have
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
\end{equation}
The explicit, optimistic-Stokes estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}
The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$, to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

As already explained, unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve climatic interactions.  On the other hand, the per-step expense is much greater because nontrivial coupled equations, indeed a free-boundary problem, must be solved for the velocity and geometry-update simultaneously.  For SIA models we parameterize the flops of such coupled solutions as $O(m^{1+\beta})$ with $\beta \ge 0$.  A large constant is assumed to be present.

@@

The only implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.  For simplified dome geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.  (An earlier non-Newton implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.  Practical steady-state ($\Delta t=\infty$) solutions at ice sheet scale are also demonstrated by \cite{JouvetBueler2012} and \cite{Bueler2016}.

>>>>

\newcommand{\nn}{\text{n}}
A diffusive description of the mass continuity equation is also valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, for the isothermal case where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}, one has $C = 2 A (\rho g)^\nn/(\nn+2)$.}  While Equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.  Generically across stress balance choices, for grounded glaciers and ice sheets one observes that large values of $d$ indicate locations of unstable mode growth if explicit time-steps are chosen too large.

@@

Table \ref{tab:notation} lists the parameters used in the numerical model performance analysis of this section.  The primary parameters are $\Delta x$, which is a representative value for the horizontal mesh (grid) cell diameter, and $m$, which is the number of nodes (vertices) in the horizontal mesh.  \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits; asymptotic and big-O notation is used only in this limit.

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution is $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry (and velocity) solution is $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit, coupled Stokes geometry and velocity solution is $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of map-plane domain & km \\
$m$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step in model years & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis.  Note $\alpha$, $\beta$, $\gamma$, and $m$ are pure, unit-less numbers.}
\label{tab:notation}
\end{table*}

Assuming the map-plane extent of the model domain is 2D and of width $L$, these primary parameters are related by
\begin{equation}
\Delta x \sim \frac{L}{m^{1/2}} \quad \text{and} \quad m \sim \frac{L^2}{\Delta x^2}.  \label{eq:paramrelation}
\end{equation}
That is, for fixed domain width $L$ there are $O(m^{1/2})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$, but this paper addresses 3D models with map-plane (2D) horizontal meshes.}

A numerical glacier or ice sheet model uses $m$ ice thickness or surface elevation variables, i.e.~one \emph{degree of freedom} per mesh node, including at ice-free nodes where the thickness has value zero.  Storing these model state variables, plus the thermodynamical state, requires $O(m)$ memory if the mesh/grid has \emph{a priori} bounded resolution in the vertical direction.  The amount of computer memory needed by the simulation is also $O(m)$, assuming prior states are discarded.

Such models also have $O(m)$ velocity variables, but note that these are not state variables.  That is, a very-viscous stress balance computes velocity as a function of the true state variables, especially ice thickness or surface elevation.

The above assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,BrinkerhoffJohnson2015,Hoffmanetal2018,Lengetal2012,
Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.  Some solvers use 3D refinement \citep{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016}, with various distinctions between how horizontal and vertical meshing is handled.  This and many other details cannot be pursued here.

Glacier and ice sheet models necessarily resolve and integrate ice-climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle, and on longer scales.  Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.  Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.  Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ describes the \emph{coupling} frequency on which ice geometry needs to be updated via solution of the mass continuity or surface kinematical equation.

Current-technology glacier and ice sheet models use explicit time-stepping which is only conditionally stable.  For the spatial resolutions used in present-day scientific applications, maintenance of explicit time-stepping stability requires steps substantially shorter than $1/q$ model years.  That is, the performance of current-generation models is limited by numerical analysis choices, and not by scientific needs.

For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996}, where $D$ is a representative diffusivity value, i.e.~of $d$ in equation \eqref{eq:siamasscontinuity}.  For Stokes or other membrane-stress-resolving dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, might be said to represent the \emph{optimistic} paradigm.  The corresponding \emph{pessimistic} paradigm requires $\Delta t < O(D^{-1} \Delta x^2)$, using a representative diffusivity value $D$ computed in the SIA manner.

Explicit time-stepping with hybrid and higher-order schemes is somewhat better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.  Some hybrid schemes apply the pessimistic paradigm as an adaptive restriction \citep{Winkelmannetal2011} across all spatial scales.  Other models apparently require the user to choose a fixed time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.  The optimistic paradigm has theoretical support for a certain higher-order DIVA scheme \citep[Equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t \approx O(\Delta x^{1.6})$ \citep[Figure 3(a)]{Robinsonetal2022}.  An intermediate power, a restriction $\Delta t = O(\Delta x^\omega)$ for some fuzzy value $1.5<\alpha<2$, also aligns with this author's experience.

Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$, but this restriction reflects the simulation purpose, not maintenance of stability.  For an implicit scheme the desired frequency of ice-climate interaction determines the total simulation cost only according to the (large) solution cost, at each time step, of solving coupled mass and momentum equations, including the surface kinematical or mass continuity equation.

@@

For each explicit time step of a model which applies a membrane-stress-resolving balance, the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  Let us assume that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.  For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 1$ if sparsity is exploited or $\alpha \approx 2$ if not \citep{Bueler2021}.  However, a multigrid method \citep{Trottenbergetal2001} can greatly reduce $\alpha$, and the ideal value $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling or solver complexity \citep{Bueler2021}.  For example, Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[Table 8.1]{IsaacStadlerGhattas2015}.  The \cite{Lengetal2012} algebraic-multigrid Stokes solver may have similar scaling, but reported results do not constrain $\alpha$.  For a higher-order stress balance on the Greenlandic ice sheet, \cite{Tuminaroetal2016} report total algebraic-multigrid-preconditioned iterations suggesting $\alpha \approx 0.05$,\footnote{See \citep[Table 7.5]{Tuminaroetal2016}.  Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.} and a geometric multigrid method by \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.  Observe that the SIA velocity computation, a trivialization of the Stokes problem in which velocity is computed by a pointwise formula, requires optimal $O(m)$ flops.

When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, and it depends strongly on solver design.  Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling.  Actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.  Also, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results, among others, all show good \emph{parallel} scaling, something not addressed here.

Regardless of the stress balance, an explicit time-stepping scheme will first compute velocity from current geometry and then apply the mass continuity equation to update the ice thickness using $O(m)$ work.  That is, once the velocity is computed from the last-known geometry, an explicit scheme replaces old thickness values by new ones using a pointwise formula.\footnote{Any additional computation needed to remesh the updated geometry, a highly design-dependent cost, is omitted here.}

@@

Now, how many flops are needed to simulate one model year?  Suppose a numerical model takes time steps of $\Delta t$ model years duration, equivalently $\Delta t^{-1}$ steps per model year.  In these terms one may write stability restrictions as required numbers of steps per model year.  That is, stability requires that $\Delta t^{-1}$ be bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.  Recalling the scaling in Equation \eqref{eq:paramrelation}, of $\Delta x$ with $m$, the explicit SIA and pessimistic-Stokes cases have
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
\end{equation}
The explicit, optimistic-Stokes estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}

The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$, to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

As already explained, unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve ice-climate interactions.  On the other hand, the per-step expense is much greater because nontrivial coupled equations, and a free-boundary problem, must be solved for the velocity and geometry-update simultaneously.  For SIA models, let us assume that the flops of such coupled solutions scale as $O(m^{1+\beta})$ with $\beta \ge 0$; a large constant is assumed.

@@

The only currently implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.  For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.  (An earlier non-Newton implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.  Practical steady-state ($\Delta t=\infty$) solutions at ice sheet scale are also demonstrated by \cite{JouvetBueler2012} and \cite{Bueler2016}.
